---
title: 'American Fisheries Society OR chapter Annual Meeting - data visualization workshop - 2023'
author: "Solene Derville - solene.derville@oregonstate.edu"
date: "Feb 28, 2023"
output:
  html_document: default
  pdf_document: default
subtitle: 'Data visualization with R and beyond: tips and tricks to publish and communicate your science'
---

### Load R libraries

```{r, echo=T, result="hide", include=T}
library(tidyverse)
library(ggpubr)
library(patchwork)
library(raster)
library(ggridges)
library(ggthemes)
library(ggrepel)
library(ggforce)
```

### Load dataset 

Extracted from GBIF: GBIF.org (09 May 2021) GBIF Occurrence Download https://doi.org/10.15468/dl.frb56w 
This file contains all GBIF occurrences of the Felidae family collected in Oregon (approximated by a box of longitude [-128.6, -119.3] and latitude [41.1, 46.2]), from 1900 to present (filtering out occurrences with uncertain coordinates)

```{r}
# import the csv in R into a dataframe. The .csv file needs to be in your working directory for this to work
# to change working directory, click on > Session > Set Working Directory > Choose Working Directory
dfel <- read.csv("dataset1_GBIF_observations.csv", sep = ";")

# this dataset has many columns, some of which are not necessary here so let's simplify the dataframe by selecting specific columns
dfel <- dfel %>% 
  dplyr::select(gbifID, family, genus, species, individualCount, decimalLatitude, decimalLongitude, eventDate, month, year, institutionCode)
```

Extracting the elevation at the position where Felidae species were observed
```{r}
# elevation downloaded from GEBCO
# this object is in raster format
r <- raster("dataset2_gebco_map.tif")

# extract elevation and add it as a column in your dataframe d
dfel$elev <- raster::extract(r, dfel[c("decimalLongitude", "decimalLatitude")])

# for the purpose of these exercises, let's simply remove the observations for which elevation was not extracted
# there are couple observations with elev < 0 that don't make sense, let's set this to 0
dfel <- dfel %>% 
  filter(!is.na(elev)) %>% 
  mutate(elev = ifelse(elev < 0, 0, elev))
```

# Practical 1: understanding ggplot

```{r}
# set up the plot. No layers are added at this stage
ggplot(dfel, aes(x = species, y = elev))

# add a layer, for example, a boxplot showing the distribution of elevations by species
ggplot(dfel, aes(x = species, y = elev)) +
  geom_boxplot()
  
# color code species, either using fill (inside of the layers) or col (contour of the layers)
ggplot(dfel, aes(x = species, y = elev)) +
  geom_boxplot(aes(fill = species))
ggplot(dfel, aes(x = species, y = elev)) +
  geom_boxplot(aes(col = species))
# note that if you try to specify col outside the aes() it doesn't work...
#ggplot(dfel, aes(x = species, y = elev)) +
  #geom_boxplot(col = species)
# unless the color is fixed and doesn't depend on a variable
ggplot(dfel, aes(x = species, y = elev)) +
  geom_boxplot(col = "royalblue")

# now let's tweak the color scale and the axis scales
ggplot(dfel, aes(x = species, y = elev)) +
  geom_boxplot(aes(col = species)) +
  scale_y_continuous("Elevation (m)", limits = c(0, 5000)) +
  scale_x_discrete("Felidae Species") +
  scale_color_canva(palette = "Playful greens and blues")

# you can create a fixed theme to use for all plots
# set my personal theme
mytheme <- theme(panel.border = element_rect(linewidth=0.5,color="black", fill="transparent"), #white background
        panel.background = element_rect(fill = 'white'),
        text=element_text(face="bold", size=8), #general size for all text.looks small here but ok once using ggsave with 600 dpi and default pointsize
        title = element_text(size=rel(1.2))) # all titles will be a little larger than labels 

# and add it to the ggplot
ggplot(dfel, aes(x = species, y = elev)) +
  geom_boxplot(aes(col = species)) +
  scale_y_continuous("Elevation (m)", limits = c(0, 5000)) +
  scale_x_discrete("Felidae Species") +
  scale_color_canva(palette = "Playful greens and blues") +
  mytheme

# the legend is redundant with the x-axis labels so we can remove it with guide = "none"
ggplot(dfel, aes(x = species, y = elev)) +
  geom_boxplot(aes(col = species)) +
  scale_y_continuous("Elevation (m)", limits = c(0, 5000)) +
  scale_x_discrete("Felidae Species") +
  scale_color_canva(palette = "Playful greens and blues", guide = "none") +
  mytheme
```


# Practical 2: patchwork

The patchwork package has a very simple syntax to make multi-panel plots and annotate them. In the following example, we will make three different plots and combine them into one figure.

```{r}
# run these lines of code to create three different plots.
g1 <- ggplot(dfel, aes(x = species, y = elev)) +
  geom_boxplot(aes(col = species)) +
  scale_y_continuous("Elevation (m)", limits = c(0, 5000)) +
  scale_x_discrete("Felidae Species")
g1

g2 <- ggplot(dfel, aes(x = decimalLongitude, y = species)) +
  # density_ridges are an alternative to boxplots
  geom_density_ridges(aes(fill = species, col = species), alpha = 0.8, scale = 0.9) +
  scale_y_discrete("Felidae Species") +
  # use parse to make the lat and long labels look good
  scale_x_continuous("Longitude", breaks = seq(-125, -119, 1),
                   labels = parse(text=paste0(seq(125, 119, -1),"^o", "*W")))
g2

g3 <- ggplot(dfel[!is.na(dfel$month),], aes(x = as.factor(month), y = elev)) +
  geom_point(aes(col = species)) +
  scale_y_continuous("Elevation (m)", limits = c(0, 5000)) +
  scale_x_discrete("Month")
g3
```

Now let's use patchwork to combine the plots together (ignore warnings, there are due to NAs in the dataset)
```{r}
# combine plots 
g <- (g1 / (g2 | g3)) # the | codes for horizontal alignment and the \ codes for vertical alignment
g

# now let's set up scales and themes for all panels at the same time using the "&" sign
g <- (g1 / (g2 | g3)) &
  scale_color_canva(guide ="none", palette = "Playful greens and blues") &
  scale_fill_canva(palette = "Playful greens and blues") & 
  mytheme
g

# label panels: a), b), c)
g <- (g1 / (g2 | g3)) &
  scale_color_canva(guide ="none", palette = "Playful greens and blues") &
  scale_fill_canva(palette = "Playful greens and blues") & 
  mytheme &
  plot_annotation(tag_levels = "A", tag_suffix = ")")
g

# move legend to the bottom
g <- (g1 / (g2 | g3)) +
  plot_layout(guides = "collect") &
  scale_color_canva(guide ="none", palette = "Playful greens and blues") &
  scale_fill_canva(palette = "Playful greens and blues") & 
  mytheme &
  plot_annotation(tag_levels = "a", tag_suffix = ")") &
  theme(legend.position = "bottom")
g
```

## Practical 3: ggforce and ggrepel

Using ggrepel to annotate elements of the plot with text
```{r}
# back to the last plot we made in the last exercise
ggplot(dfel[!is.na(dfel$month),], aes(x = as.factor(month), y = elev)) +
  geom_point(aes(col = species)) +
  scale_y_continuous("Elevation (m)") +
  scale_x_discrete("Month")


# the purpose of this plot is to check if there is a pattern in elevation as a function of season
# there are clearly two outliers showing elevation > 2250 m. Let's use ggrepel to annotate those points on the plot
# for example, we could label those points with the name of the source that provided the record and the year of the record
ggplot(subset(dfel, !is.na(dfel$month)), aes(x = as.factor(month), y = elev)) +
  geom_point(aes(col = species)) +
  scale_y_continuous("Elevation (m)") +
  scale_x_discrete("Month") +
  geom_text_repel(data = subset(dfel, elev > 2250 & !is.na(dfel$month)), 
                  aes(label = paste(institutionCode, year),
                      col = species), 
                  box.padding = 0.8, max.overlaps = Inf)

# run it several times, the plot will change
```

Using ggforce to draw ellipses, rectangles, circles etc around elements of the plot
```{r}
# make a very basic plot of longitude and latitude of puma observations colored by elevation
dfel_g3 <- dfel %>% 
  filter(!is.na(dfel$month) & species == 'Puma concolor')
g3 <- ggplot(data = dfel_g3,
       aes(decimalLongitude, decimalLatitude)) +
    geom_point(aes(col = elev)) +
    xlab("Longitude") +
    ylab("Latitude") +
    coord_fixed()
g3

# now let's circle observations made at high altitude
# the geom_mark_ellipse function has a convenient "filter" argument to select just a subset of the data, here the records at elevation > 1500 m
g3 + geom_mark_rect(aes(filter = elev > 1500))

# complement the function to add label these points
# the dataframe needs to hold the labels and descriptions into a column
dfel_g3 <- dfel_g3 %>% 
  mutate(desc = ifelse(elev > 1500, "Alt > 1500 m", ""))

#Observations in Oregon at high altitude > 1500 m
ggplot(data = dfel_g3,
       aes(decimalLongitude, decimalLatitude)) +
    geom_point(aes(col = elev)) +
    xlab("Longitude") +
    ylab("Latitude") +
    coord_fixed() +
    geom_mark_ellipse(aes(filter = elev > 1500,
                           label = species,
                           description = desc))
# it is possible that the label of the ellipse will not plot properly if your plot window is too small
# if that is the case, click on "zoom" in your plot panel
```


# Practical 4: adding stats


```{r}
# make that first plot again and store it in object g_final
g4 <- ggplot(dfel, aes(x = species, y = elev)) +
  geom_boxplot(aes(col = species)) +
  scale_y_continuous("Elevation (m)", limits = c(0, 5000)) +
  scale_x_discrete("Felidae Species") +
  scale_color_canva(palette = "Playful greens and blues", guide = "none") +
  mytheme
g4

# the data does not seem to follow a normal distribution so we will use a kruskal wallis test to compare elevation across species
# set levels of comparison for kruskal wallis test  
my_comparisons <- list( c("Felis catus", "Lynx rufus"), c("Lynx rufus", "Puma concolor"), c("Felis catus", "Puma concolor"))

g4 <- g4 +
  stat_compare_means(comparisons = my_comparisons, 
                     label.y = c(2000, 2800, 3500), # specify where on the plot you want the test results to appear
                     label = "p.signif", 
                     p.adjust.method = "none") + # Add pairwise comparisons p-value
  stat_compare_means(label.y = 4500)    # Add global p-value
g4
```

# Practical 5: saving


Save in different formats, files types and resolutions
```{r}
# save in a low resolution jpeg, then zoom on the image, it should be very pixelized
ggsave(g4, file = "myfigure_badresolution.jpeg", width = 88*2, height = 88*2, dpi = 90, units = "mm")

# save in a high resolution tiff, full page width format
ggsave(g4, file = "myfigure_highquality.tiff", width = 88*2, height = 88*2, dpi = 600, units = "mm")

# save in a vector image format: pdf. No resolution in dpi is required
ggsave(g4, file = "myfigure_highquality_vector.pdf", width = 88*2, height = 88*2, units = "mm")
# you can zoom as much as you want and the lines remain very clear. 
# also, compared to tiff, the file is much much smaller!

# save as png with a transparent background
ggsave(g4, file = "myfigure_highquality_transparent.png", width = 88*2, height = 88*2, dpi = 600, units = "mm", bg = "transparent")
```


